### 2023-10-30 journal/algorithm-GC.tex
editing space bounded version

\section{Detailed Description of the Bounded-Space Implementation}
\label{reducing-details}

@@

Here, we give more details of the bounded-space construction sketched in Section \ref{reducing}.
We first describe the modifications to the implementation.
To avoid confusion, we use nodes to refer to the nodes of the ordering tree, and blocks to refer
to the nodes of a RBT (since the RBT stores blocks).

The space-bounded implementation uses two additional shared arrays:
the \var{leaf} array allows processes to access one another's leaves to perform helping, and
the \var{last} array is used to determine which blocks are safe to discard.
\begin{itemize}
\item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ is the largest \fld{index} of a block in the}\\
\mbox{ }\Comment{root that process $k$ has observed to contain either an en-}\\
\mbox{ }\Comment{queue whose value has been dequeued or a null dequeue }
\end{itemize}

@@

Blocks no longer require the \fld{super} field.  It was used to quickly find a block's 
superblock in the parent node's \fld{blocks} array, but this can now be done efficiently 
by searching the parent's \fld{blocks} RBT instead.
Each \typ{Block} has an additional field.
\begin{itemize}
\item \typ{int} \var{index} \Comment{position this block would have in the \fld{blocks} array}
\end{itemize}
To facilitate helping, each \typ{Block} in a leaf has one more additional field that is used only for blocks that store a dequeue operation.
\begin{itemize}
\item \typ{Object} \var{response} \Comment{response of the dequeue in the block}
\end{itemize}

\here{Talk about what happens if a block is not found in an RBT}

@@

\begin{figure*}
\begin{algorithmic}[1]
\setcounter{ALG@line}{200}

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
    \gc{
    \State \var{h} \assign\ \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}+1
	\State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \var{e},
		\fld{sum\sub{enq}}\ \assign\ $\var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{enq}}+1,$\\
		\fld{sum\sub{deq}}\ \assign\ $\var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{deq}}$,
		\fld{index}\ \assign $h$}\label{enqNewGC}}
    \State \Call{Append}{\var{B}}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
    \gc{
    \State \var{h} \assign\ \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}+1
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
	    \fld{sum\sub{enq}}\ \assign\ $\var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{enq}},$\\
	    \fld{sum\sub{deq}}\ \assign\ $\var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{deq}}+1$,
	    \fld{index} \assign $h$}\label{deqNewGC}
	}
    \State \Call{Append}{\var{B}}
    \gc{\State \Return \Call{CompleteDeq}{\var{leaf}, $h$}}
\EndFunction{Dequeue}

@@

\gc{
\Function{Object}{CompleteDeq}{\var{leaf}, $h$} \Comment{finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ $\Call{IndexDequeue}{\var{leaf}, h, 1}$
    \State \var{res} \assign $\Call{FindResponse}{b, i}$
	\State \Return{\var{res}}
\EndFunction{CompleteDeq}
}

@@

\Function{void}{Append}{\typ{Block} \var{B}}  \Comment{append block to leaf and propagate to root}
    \gc{\State \var{leaf}.\fld{blocks} \assign\ \Call{AddBlock}{\var{leaf}, \var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
\EndFunction{Append}

@@

\gc{
\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}} \Comment{add block $\var{B}\neq\nl$ to \var{T}; do GC if necessary}
    \If{\var{B}.\fld{index} is a multiple of $G$} \Comment{do garbage collection}\label{GCstart}
        \State $s$ \assign\ \Call{SplitBlock}{\var{v}}.\var{index}\label{callSplitBlock}
        \State \Call{Help}{}\label{invokeHelp}
        \State $T'$ \assign\ \Call{Split}{$T,s$} \Comment{\op{Split} removes blocks with $\fld{index} < s$}\label{splitGC}
        \State \Return \Call{Insert}{$T',B$}\label{GCend}
	\Else\ \Return \Call{Insert}{$T, B$}
    \EndIf
\EndFunction{AddBlock}
}

\end{algorithmic}
\caption{Bounded-space queue implementation's main routines.\label{code1GC}}
\end{figure*}

\begin{figure*}
\begin{algorithmic}[1]
\setcounter{ALG@line}{247}
\gc{
\Function{Block}{SplitBlock}{\typ{Node} v} \Comment{figure out where to split \var{v}'s RBT}
	\If{$\var{v}=\var{root}$}
		\State $m$ \assign\ 0
		\For{$k$ \assign\ $1..p$} $m$ \assign\ $\max(m, \var{v}.\fld{last}[k])$
		\EndFor
		\State $B$ \assign\ $\var{root}.\fld{blocks}[m-1]$\label{SBlookup1}
	\Else
		\State $B_p$ \assign\ $\Call{SplitBlock}{\var{v}.\fld{parent}}$\label{SBrecurse}
		\State \var{dir} \assign\ ($\var{v}=\var{v}.\fld{parent.left}$ ? \fld{left} : \fld{right})
		\State $B$ \assign\ $\var{v}.\fld{blocks}[B_p.\edir]$\label{SBlookup2}
	\EndIf
	\State \Return ($B = \nl$ ? $\Call{MinBlock}{\var{v}.\var{blocks}}$ : $B$) \Comment{If $B$ was discarded, use leftmost block instead}\label{substitute}
\EndFunction{SplitBlock}
}

@@

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \linecomment{create new block $B$ to install in \var{v}.\fld{blocks}[\var{i}]}
    \State \hangbox{let \var{B} be a new \typ{Block} with fields\label{initNewBlockGC}
    \gc{
    	\eleft\ \assign\ $\Call{MaxBlock}{\var{v}.\fld{left}.\fld{blocks}}.\fld{index}$\label{createEndLeftGC},\\
    \eright\ \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$,\label{createEndRightGC}
    \fld{index} \assign\ $i$\label{createIndexGC}
    }\\
	\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{enq}} + 
			\var{v}.\fld{right.blocks}[\var{B}.\eright].\fld{sum\sub{enq}}\\
	\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{deq}} +
			\var{v}.\fld{right.blocks}[\var{B}.\eright].\fld{sum\sub{deq}}}
    \State \var{num\sub{enq}} \assign\ $\var{B}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
    \State \var{num\sub{deq}} \assign\ $\var{B}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$} \var{B}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + 
        	\var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLengthGC}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl \hfill \Comment{no blocks need to be propagated to \var{v}}\label{testEmptyGC}\label{CBnullGC}
    \Else\ \Return \var{B}
    \EndIf
\EndFunction{CreateBlock}

@@

\begin{figure*}
\begin{algorithmic}[1]
\setcounter{ALG@line}{290}
\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated to root and $|D(\var{v}.\fld{blocks}[\var{b}])|\geq \var{i}$}
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
    \Else
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\gc{
		\State $T$ \assign\ $\var{v}.\fld{parent.blocks}$
		\State $B_p$ \assign\ min block in $T$ with $\edir \geq b$\label{searchsuper1}
		\State $B_p'$ \assign\ max block in $T$ with $\edir < b$\label{searchsuper2}
	    }
	    \State \linecomment{compute index \var{i} of dequeue in superblock $B_p$}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -
	    		\var{v}.\fld{blocks}[\gc{B_p'}.\edir].\fld{sum\sub{deq}}$}
        \If{$\fld{dir} = \fld{right}$} 
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\gc{B_p}.\eleft].\fld{sum\sub{deq}} - \mbox{ }
					\var{v}.\fld{blocks}[\gc{B_p'}.\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
        \EndIf \label{computeISuperEndGC}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, $\gc{B_p.\fld{index}}$, \var{i}}
    \EndIf
\EndFunction{IndexDequeue}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}} \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} -
    		\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$\label{FRnumenqGC}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC} \Comment{queue is empty when dequeue occurs}
        \gc{
        \If{$b > \var{last}[\var{id}]$} $\var{last}[\var{id}]$\ \assign\ $b$
        \EndIf
        }
        \State \Return \nl \hfill \label{returnNullGC}
    \Else \ \Comment{response is the \var{e}th enqueue in the root}
        \State $\var{e}$\ \assign\ $\var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size$\label{computeEGC}

@@

The new routines, \op{AddBlock}, \op{SplitBlock}, \op{Help} and \op{Propagated} are
used to implement the garbage collection (GC) phase.
When a \op{Refresh} or \op{Append} wants to add a new block to a node's \fld{blocks} RBT,
it calls the new \opemph{AddBlock} routine.
Before attempting to add the block to a node's RBT, \op{AddBlock} triggers a GC phase on the RBT if 
the new block's \fld{index} is a multiple of the constant $G$, which we choose to be $p^2\ceil{\log p}$.
This ensures that obsolete blocks are removed from the RBT once every $G$ times a new block is added to it.
The GC phase uses \op{SplitBlock} to determine the index $s$ of the oldest block to keep,
calls \op{Help} to help all pending dequeues that have been propagated to the root
(to ensure that all blocks before $s$ can safely be discarded),
uses the standard RBT \op{Split} routine \cite{Tar83}
to remove all blocks with \fld{index} less than $s$,
inserts the new block
and finally performs a \op{CAS} to swing the node's \fld{blocks} pointer to the new RBT.

To determine the oldest block in a node \var{v} to keep, the \opemph{SplitBlock} routine  first
uses the \var{last} array to
find the most recent block $B_{root}$ in the root that contains either an enqueue that has been dequeued
or a null dequeue.  By the FIFO property of queues, all enqueues in blocks before $B_{root}$ are either
dequeued or will be dequeued by a dequeue that is currently in progress.  Once those pending dequeues have been
helped to complete by line \ref{invokeHelp},
it is safe to discard any blocks in the root older than $B_{root}$, as well
as their subblocks.\footnote{If we used the more conservative approach of discarding blocks whose indices 
are smaller than the \emph{minimum} entry of \var{last} instead of the maximum, helping would be unnecessary, but then one slow process could prevent GC from discarding any blocks, so the space would not be bounded.}
The \op{SplitBlock} uses the \eleft\ and \eright\ fields to find the last block in \var{v} that is a subblock
of $B_{root}$ (or any older block in the root, in case $B_{root}$ has no subblocks in \var{v}).
As \op{SplitBlock} is in progress, it is possible that some block that it needs in a node $\var{v}'$
along the path from \var{v} to the root is discarded by another GC phase.
In this case, \op{SplitBlock} uses the last subblock in \var{v} of the oldest block in $\var{v}'$ instead
(since a GC phase on $\var{v}'$ determined that all blocks older than that are safe to discard anyway).

@@

In the code, we use $\var{v}.\fld{blocks}[i]$ to refer to the block in the RBT stored in $\var{v}.\fld{blocks}$
with index~$i$.
A search for this block may sometimes not find it, if it has already been discarded
by another process's GC phase.
As mentioned in Section \ref{reducing}, if this happens to an enqueue operation,
the enqueue can simply terminate because the fact that the block is gone means that another
process has helped the enqueue reach the root of the ordering tree.
Similarly, if  a dequeue operation performs a failed search on a RBT, the dequeue can return the value 
written in the \fld{response} field of the 
leaf block that represents the dequeue and terminate, since some other process
will have written the \fld{response} there before discarding the needed block.
We do not explicitly write this early termination in the pseudocode every time we do a lookup in an RBT.
There is one exception to this rule:  if an RBT lookup for block $B$ returns \nl\
on line \ref{SBlookup1} or \ref{SBlookup2} of
\op{SplitBlock} because the required block has been discarded, 
we continue doing GC, since we do not want a GC phase on one node
to be prevented from cleaning up its RBT because a GC phase on a different node threw away some blocks
that were needed.
Line \ref{substitute} says what to do in this case.

>>>>

\subsection{Detailed Description}
\label{reducing-details}

@@

To avoid confusion, we use nodes to refer to the nodes of the ordering tree, and blocks to refer
to the nodes of a RBT (since the RBT stores blocks).
The space-bounded implementation uses two  shared arrays:
the \var{leaf} array allows processes to access one another's leaves to perform helping, and
the \var{last} array is used to determine which blocks are safe to discard.

@@

Blocks no longer require the \fld{super} field.  It was used to quickly find a block's 
superblock in the parent node's \fld{blocks} array, but this can now be done efficiently 
by searching the parent's \fld{blocks} RBT instead.
Each \typ{Block} has an additional field.
\begin{itemize}
\item \typ{int} \var{index} 
\end{itemize}
that represents the position this block would have in the \fld{blocks} array.
To facilitate helping, each \typ{Block} in a leaf has one more additional field 
\begin{itemize}
\item \typ{Object} \var{response}
\end{itemize}
which is used only for blocks that store a dequeue operation and stores the response of the dequeue in the block.

@@

\begin{figure*}
\begin{algorithmic}[1]
\setcounter{ALG@line}{200}
\State \linecomment Shared variables
\State \typ{Node} root \Comment{root of ordering tree}
\gc{
\State \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
\State \typ{int[]} \var{last}$[1..p]$\Comment{\var{last}$[k]$ is max \fld{index} of a root block that process $k$ saw}
\State \mbox{ }\Comment{contains either a dequeued enqueue or a null dequeue}
}

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}}

    \gc{\State \var{h} \assign\ \Call{MaxBlock}{\var{leaf}[\var{id}].\fld{blocks}}.\fld{index}+1}
	\State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \var{e},
		\fld{sum\sub{enq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{enq}}+1,$\\
		\fld{sum\sub{deq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{deq}}$,
		\gc{\fld{index}\ \assign $h$}}\label{enqNewGC}
    \State \Call{Append}{\var{B}}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
    \State \gc{\var{h} \assign\ \Call{MaxBlock}{\var{leaf}[\var{id}].\fld{blocks}}.\fld{index}+1}
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
	    \fld{sum\sub{enq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{enq}},$\\
	    \fld{sum\sub{deq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{deq}}+1$,
	    \gc{\fld{index} \assign $h$}}\label{deqNewGC}
    \State \Call{Append}{\var{B}}
    \gc{\State \Return \Call{CompleteDeq}{\var{leaf}[\var{id}], $h$}}
\EndFunction{Dequeue}

@@

\gc{
\Function{Object}{CompleteDeq}{\typ{Node} \var{leaf}, \typ{int} $h$} \Comment{finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ $\Call{IndexDequeue}{\var{leaf}, h, 1}$
    \State \Return $\Call{FindResponse}{b, i}$
\EndFunction{CompleteDeq}
}

\Function{void}{Append}{\typ{Block} \var{B}}  \Comment{append block to leaf and propagate to root}
    \gc{\State \var{leaf}[\var{id}].\fld{blocks} \assign\ \Call{AddBlock}{\var{leaf}[\var{id}], \var{leaf}[\var{id}].\fld{blocks}, \var{B}}\label{appendLeafGC}}
    \State \Call{Propagate}{\var{leaf}[\var{id}].\fld{parent}} 
\EndFunction{Append}

@@

\gc{
\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}} \Comment{add block $\var{B}\neq\nl$ to \var{T}; do GC if necessary}
    \If{\var{B}.\fld{index} is a multiple of $G$} \Comment{do garbage collection}\label{GCstart}
        \State $s$ \assign\ \Call{SplitBlock}{\var{v}}.\var{index}\label{callSplitBlock}
        \State \Call{Help}{}\label{invokeHelp}
        \State $T'$ \assign\ \Call{Split}{$T,s$} \Comment{\op{Split} removes blocks with $\fld{index} < s$}\label{splitGC}
        \State \Return \Call{Insert}{$T',B$}\label{GCend}
	\Else\ \Return \Call{Insert}{$T, B$}\label{noGC-insert}
    \EndIf
\EndFunction{AddBlock}
}

\end{algorithmic}
\caption{Bounded-space queue implementation's main routines for process number $id$.\label{code1GC}}
\end{figure*}

\begin{figure*}
\begin{algorithmic}[1]
\setcounter{ALG@line}{251}
\gc{
\Function{Block}{SplitBlock}{\typ{Node} v} \Comment{figure out where to split \var{v}'s RBT}
	\If{$\var{v}=\var{root}$}
		\State $m$ \assign\ 0
		\For{$k$ \assign\ $1..p$} $m$ \assign\ $\max(m, \var{v}.\fld{last}[k])$
		\EndFor
		\State $B$ \assign\ $\var{root}.\fld{blocks}[m-1]$\label{SBlookup1}
	\Else
		\State $B_p$ \assign\ $\Call{SplitBlock}{\var{v}.\fld{parent}}$\label{SBrecurse}
		\State \var{dir} \assign\ ($\var{v}=\var{v}.\fld{parent.left}$ ? \fld{left} : \fld{right})
		\State $B$ \assign\ $\var{v}.\fld{blocks}[B_p.\edir]$\label{SBlookup2}
	\EndIf
	\State \Return ($B = \nl$ ? $\Call{MinBlock}{\var{v}.\var{blocks}}$ : $B$) \Comment{If $B$ was discarded, use leftmost block instead}\label{substitute}
\EndFunction{SplitBlock}
}

@@

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \linecomment{create new block $B$ to install in \var{v}.\fld{blocks}[\var{i}]}
    \State \hangbox{let \var{B} be a new \typ{Block} with fields\label{initNewBlockGC}
    \gc{
    	\eleft\ \assign\ $\Call{MaxBlock}{\var{v}.\fld{left}.\fld{blocks}}.\fld{index}$\label{createEndLeftGC},\\
    \eright\ \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$,\label{createEndRightGC}
    \fld{index} \assign\ $i$\label{createIndexGC}
    }\\
	\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{enq}} + 
			\var{v}.\fld{right.blocks}[\var{B}.\eright].\fld{sum\sub{enq}}\\
	\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{deq}} +
			\var{v}.\fld{right.blocks}[\var{B}.\eright].\fld{sum\sub{deq}}}
    \State \var{num\sub{enq}} \assign\ $\var{B}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
    \State \var{num\sub{deq}} \assign\ $\var{B}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$} \var{B}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + 
        	\var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLengthGC}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl\ \Comment{no blocks need to be propagated to \var{v}}\label{testEmptyGC}\label{CBnullGC}
    \Else\ \Return \var{B}
    \EndIf
\EndFunction{CreateBlock}

@@

\begin{figure*}
\begin{algorithmic}[1]
\setcounter{ALG@line}{296}
\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{return $\langle b', i' \rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is ($i'$)th dequeue of $D(\var{root}.\fld{blocks}[b'])$}
    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated to root and $|D(\var{v}.\fld{blocks}[\var{b}])|\geq \var{i}$}
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
    \Else \Comment{First, find the superblock $B_p$ of $v.blocks[b]$}
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\gc{
		\State $T$ \assign\ $\var{v}.\fld{parent.blocks}$
		\State $B_p$ \assign\ min block in $T$ with $\edir \geq b$\label{searchsuper1}
		\State $B_p'$ \assign\ max block in $T$ with $\edir < b$ \Comment{predecessor of $B_p$}\label{searchsuper2}}
	    \State \linecomment{compute index \var{i} of dequeue in superblock $B_p$}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -
	    		\var{v}.\fld{blocks}[\gc{B_p'}.\edir].\fld{sum\sub{deq}}$}
        \If{$\fld{dir} = \fld{right}$} 
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\gc{B_p}.\eleft].\fld{sum\sub{deq}} - \mbox{ }
					\var{v}.\fld{blocks}[\gc{B_p'}.\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
        \EndIf \label{computeISuperEndGC}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, $\gc{B_p.\fld{index}}$, \var{i}}
    \EndIf
\EndFunction{IndexDequeue}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}} \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} -
    		\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$\label{FRnumenqGC}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC} \Comment{queue is empty when dequeue occurs}%
        \gc{\If{$b > \var{last}[\var{id}]$} $\var{last}[\var{id}]$\ \assign\ $b$
        \EndIf
        }
        \State \Return \nl \hfill \label{returnNullGC}
    \Else \ \Comment{response is the \var{e}th enqueue in the root}
        \State $\var{e}$\ \assign\ $\var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size$\label{computeEGC}

@@

The new routines, \op{AddBlock}, \op{SplitBlock}, \op{Help} and \op{Propagated} are
used to implement the garbage collection (GC) phase.
When a \op{Refresh} or \op{Append} wants to add a new block to a node's \fld{blocks} RBT,
it calls the new \opemph{AddBlock} routine.
Before attempting to add the block to a node's RBT, \op{AddBlock} triggers a GC phase on the RBT if 
the new block's \fld{index} is a multiple of the constant $G$, which we choose to be $p^2\ceil{\log p}$.
This ensures that obsolete blocks are removed from the RBT once every $G$ times a new block is added to it.
The GC phase uses \op{SplitBlock} to determine the index $s$ of the oldest block to keep,
calls \op{Help} to help all pending dequeues that have been propagated to the root
(to ensure that all blocks before $s$ can safely be discarded), and then
uses the standard RBT \op{Split} routine \cite{Tar83}
to remove all blocks with \fld{index} less than $s$.
Then \op{AddBlock}
inserts the new block at line \ref{GCend} or \ref{noGC-insert}
and returns the root of the resulting RBT.
The \op{Append} or \op{Refresh} then stores this root into the node's \fld{blocks} field at line 
 \ref{appendLeafGC} or \ref{casGC}.

To determine the oldest block in a node \var{v} to keep, the \opemph{SplitBlock} routine  first
uses the \var{last} array to
find the most recent block $B_{root}$ in the root that contains either an enqueue that has been dequeued
or a null dequeue.  By the FIFO property of queues, all enqueues in blocks before $B_{root}$ are either
dequeued or will be dequeued by a dequeue that is currently in progress.  Once those pending dequeues have been
helped to complete by line \ref{invokeHelp},
it is safe to discard any blocks in the root older than $B_{root}$, as well
as their subblocks.\footnote{If we used the more conservative approach of discarding blocks whose indices 
are smaller than the \emph{minimum} entry of \var{last} instead of the maximum, helping would be unnecessary, but then one slow process could prevent GC from discarding any blocks, so the space would not be bounded.}
The \op{SplitBlock} uses the \eleft\ and \eright\ fields to find the last block in \var{v} that is a subblock
of $B_{root}$ (or any older block in the root, in case $B_{root}$ has no subblocks in \var{v}).
While \op{SplitBlock} is in progress, it is possible that some block that it needs in a node $\var{v}'$
along the path from \var{v} to the root is discarded by another GC phase.
In this case, \op{SplitBlock} uses the last subblock in \var{v} of the oldest block in $\var{v}'$ instead
(since a GC phase on $\var{v}'$ determined that all blocks older than that are safe to discard anyway).

@@

In the code, we use $\var{v}.\fld{blocks}[i]$ to refer to the block in the RBT stored in $\var{v}.\fld{blocks}$
with index~$i$.
A search for this block may sometimes not find it, if it has already been discarded
by another process's GC phase.
As mentioned above, if this happens to an enqueue operation,
the enqueue can simply terminate because the fact that the block is gone means that another
process has helped the enqueue reach the root of the ordering tree.
Similarly, if  a dequeue operation performs a failed search on a RBT, the dequeue can return the value 
written in the \fld{response} field of the 
leaf block that represents the dequeue and terminate, since some other process
will have written the \fld{response} there before discarding the needed block.
We do not explicitly write this early termination in the pseudocode every time we do a lookup in an RBT.
There is one exception to this rule:  if an RBT lookup for block $B$ returns \nl\
on line \ref{SBlookup1} or \ref{SBlookup2} of
\op{SplitBlock} because the required block has been discarded, 
we continue doing GC, since we do not want a GC phase on one node
to be prevented from cleaning up its RBT because a GC phase on a different node threw away some blocks
that were needed.
Line \ref{substitute} says what to do in this case.

\er{Edited this far}
